\label[attached-files]
\app Attached files

\app Prompts
\sec People v1

\begtt
Extract all people mentioned in the Czech newspaper articles (the article 
release timestamp is included) and output the result in JSON.

The order of people is given by their appearance in the articles.
 - Person "id" is a unique integer assigned to each person
 (start with 1).
 - A person's "name" is a unique name of the person.
   - Use full names (if available in the article) and state them 
   in a basic form (e.g., "Andrej Babiš" instead of "Andreje Babiše").
   - Use the Czech "ml." and "st." abbreviations for junior and senior 
   if stated in the text.
 - Replace each item of the "roles" array with a short Czech string 
 describing the role of the person
   - Examples: "prezident", "novinář", "předseda vlády", "soudce", etc.
   - The roles must not repeat, i.e., the "roles" array corresponds 
   to a set.
   - The roles must be atomic. An item like "premiér a prezident" 
   must be split into two roles.
   - Be as specific as possible (in line with the article text) 
   in the role description. E.g., if a person is a colleague 
   of some other person, make it clear.
   - Prefer abbreviations, e.g., "ČR" instead of "Česká republika".
 - You will receive a list of global people IDs, names and roles. 
 If you encounter a person with the same name and role, 
 you must use the same ID. If the person is not in the list, 
 assign a new ID, while extending the global list of people.
 - The IDs must be unique across all articles. If a person appears 
 in multiple articles, use the same ID.

IMPORTANT:
   - First extract the global list of people from the articles 
   and assign them IDs.
   - Use the initial IDs from the provided global list of people.
   - Then extract the list of articles and assign IDs to them.
   - Do not use "\\u" escape sequences in JSON strings. You are 
   not limited to ASCII.
   - Do not make things out - each person and his/her roles 
   must be clearly mentioned in the text.
   - Return only the JSON. Give no additional text! 
   Start with the "{" character and end with "}".
   - The person has to be unique across all of the articles
   - No unique person can appear more then one unique ID
   - The output has to start with the "{" character and end with "}".

   
Requested JSON Schema:
{{ output_schema }}

Global list of people:
{{ people }}

Czech text of the articles follow:
{{#articles}}
<article>
    ID: {{ id }}
    TEXT: {{ ftext }}
</article>
{{/articles}}

Do not output anything else than the JSON with Requested JSON Schema. 
Do not add any additional text or explanation. Do not output 
the reasoning process.
\endtt